An antibody-escape calculator for mutations to the SARS-CoV-2 receptor-binding domain

### 2022-03-06 paper/reviewer_response.tex

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb}
\usepackage{hyperref} 
\usepackage[round,semicolon]{natbib}

\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}
\newcommand{\response}[1]{{\color{black}#1}}

\title{Response to reviews of ``An antibody-escape calculator for mutations to the SARS-CoV-2 receptor-binding domain'' for \textit{Virus Evolution}}
\author{}

\begin{document}
\maketitle

\emph{Below, the reviewer and editor comments {\color{blue} are in blue}, and may responses are in black.}

\color{blue}

\subsection*{Associate Editor: Adi Stern}

Apologies for the delayed response, reviewing during pandemic seems to have become tricky.
Both reviewers agree this is an excellent paper and a valuable resource.
Both suggest a series of slight changes and clarifications that all seem valuable and doable.
I agree with reviewer 2 that perhaps``estimator'' might be a better term than calculator to emphasize uncertainty. 
Finally, I agree with the authors discussion (and reviewer 2) on the caveat of these data being based on pre-variant escape data.
I do wonder how difficult it will be to implement newer versions that are based on future data with more complex variants?
Perhaps worth adding a few words on this as well to the discussion.

\response{Thanks to you and the reviewers for the helpful reviews. In the revised version, we have addressed all of the specific suggestions, including the three you mention above (see below for details).}

\subsection*{Reviewer: 1}

One of the key objectives for SARS-CoV-2 surveillance is to rapidly predict the impact that amino acid substitutions would have on important virus characteristics, including how well the spike receptor binding domain (RBD) will be recognized by therapeutic monoclonal antibodies and by polyclonal sera from subject following vaccination or infection.  The Bloom group has previously published a series of papers in which deep mutational scanning has been used to introduce all 20 amino acid substitutions at every site within the spike RBD and the impact on binding of monoclonal antibody and polyclonal antisera measured experimentally.  In this paper, Greaney et al. use this experimental data to develop an antibody-escape calculator to predict the functional impact of any combination of substitutions in the RBD on antibody binding.  They used data from three publications that measured virus neutralization of pseudoviruses carrying specific amino acid substitutions against convalescent or vaccine sera.  This calculator will be extremely helpful for the evaluation of emerging SARS-CoV-2 variants.

\comment{Thanks for the nice summary of the work.}

While the significance of the work is very high, the manuscript would be improved by addressing the following points.

For every monoclonal antibody, a single escape fraction value is calculated at each site by summarizing the impact of all 20 amino acid substitutions at the site as either the mean impact or the sum of the impacts to produce their escape maps.  When assessing the impact of a new mutation on binding escape, any mutation at that site appears to be treated equivalently.  An alternative approach would be to just use the specific escape fraction for the specific amino acid substitution at that site.  Thus, E484K would be treated differently from E484Q and E484A.  Indeed, one would expect that some substitutions would be more or less tolerated in terms of antibody binding.  The authors may have already evaluated this approach.  At a minimum, this point should be discussed in the manuscript.  But it might also be worthwhile to compare the performance of this more specific approach to the more generic approach implemented here.

It is a little difficult to know exactly how the different parameters are being calculated since in many cases they are just describe in the text.  For example, in the Methods section they state that “We also normalize the site-level escape metrics for each antibody to account for difference in strengths of antibody selection” and that “essentially this corresponds to scaling the escape value for each antibody so that a value of one corresponds to the larger of the maximum escape as a site or 20 times the median value across sites”.  It would be much easier to understand how these values are being calculated if the actual equations for each step in the workflow were provided.

There are several ways in which the binding score could be calculated, e.g., mean versus sum, specific substitutions versus generic substitutions, etc., but the authors only present the results from one approach.  It would be useful to present results from multiple configurations to determine if the results are robust to these configuration parameters.

The most important aspect of the paper is the correlation between the calculated binding score and the experimentally-measure fold change in virus neutralization as presented in Figure 3, which would provide some kind of validation of the computational method.  The authors state in the text that “the escape-calculator scores correlate quite well with the measured neutralization titers across all studies and cohorts”; “quite well” seems pretty subjective.  At a minimum, the strength of correlation should be quantified by using something like a Spearman’s rho metric which could then be used to calculate a t test statistic and p-value.  Indeed, these metrics could be used to compare the performance of the different configuration parameters.

It is not clear how to use the online calculator.  The instructions online refer to a Tweet chain but that provides a summary of how the calculations are performed, not on how to use the online tool.  It is not very intuitive and could use some simple explicit instructions on the page itself.

\subsection*{Reviewer: 2}

Greaney et al have developed a simple tool for estimating the impact of amino acid changes on antibody protection from SAR-CoV-2 variants. This is based on experimental data, much of which has been generated by these researchers. The text is clear and well written, the method is both sound and an easy to use implementation is provided. I only have some minor comments:

Please make it clear in the abstract that the experimental data is generated on the early pre-variant of concern data and so mutations may differ in consequence on new variants such as Omicron. This would seem to be quite a big caveat with this type of data.

Line 30, variants with reduced Ab neutralisation properties emerged before 2 years, e.g, Beta was designated in late 2020 in South Africa.

Line 41, briefly define variants of concern as there's ongoing confusion between these and just variants.

Line 54, the pop here at "complex black-box computational methods" is unwarranted and out of date relative to much contemporary machine leaning methods. Arguably simple residue based information from experiments is also a black-box as the mechanistic nuance of three-dimensional structural interactions are not really being accounted for. Maybe the authors could comment in the Discussion on why such a simple method works so well, e.g., linked to immunodominance of this spike region.

Two lines into Results, mutations strictly speaking occur at the nucleotide level and result in amino acid replacements. Clearly the authors know this and it fine to use mutations as short-hand for the residue changes but at least make the distinction clear the first time this is referred to.

Results, 3rd and 4th paragraph, let's not 'imagine' anything, please re-write.

Page 2, "illustrative" would be more accurate than "toy" example.

Finally, the authors should reflect on whether this is a "calculator" as opposed to an antibody-escape "estimator", i.e., capture the inaccuracy in the inferences a bit more clearly.

\response{This is a good suggestion, and we have changed the text to use ``estimator'' rather than ``calculator.''}

\color{black}
\bibliographystyle{mbe}
{\small
\bibliography{references.bib}
}

\end{document}

### 2022-03-06 paper/paper_original_submission.tex
set up to address reviewer comments

\lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have emerged only $\sim$2 years since the virus first started to spread in humans.
A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{wang2021antibody,uriu2021neutralization,lucas2021impact}.
Unfortunately, the rate at which new variants arise outstrips the speed with which these experiments can be performed.

@@

\subsection{The escape calculations correlate well with neutralization assays of human polyclonal sera against SARS-CoV-2 variants}
For each set of RBD mutations, we can define a quantitative score that represents the polyclonal antibody binding that remains after making these mutations.
This score is defined using the same principle as the site-wise escape calculator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
This calculation is implemented in the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).

\begin{SCfigure*}[][]
\includegraphics[width=0.7\textwidth]{../results/neut_studies/neut_studies.png}
\caption{Correlation of calculated binding scores with experimentally measured fold changes in neutralization for SARS-CoV-2 variants and mutants.
The data of \citet{lucas2021impact} was generated using live virus and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
The data of \citet{uriu2021neutralization} and \citet{wang2021antibody} was generated using pseudovirus against convalescent (top) or vaccine (bottom) sera, with vaccine sera from Pfizer BNT162b2 or Moderna mRNA-1273 vaccines, respectively. 
The fold changes represent the geometric means over all subjects in each cohort.
An interactive version of this figure that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/neut_studies.html}.
}
\label{fig:neut_studies}
\end{SCfigure*}

@@

\section{Discussion}
We have described an escape calculator that uses experimental data for 33 monoclonal antibodies to estimate the antigenic effects of arbitrary combinations of mutations to the SARS-CoV-2 RBD.
The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus synergistic effects on antigenic escape.
For instance, sites 417, 484, and 490 are all peaks of antibody escape.
But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have synergistic effects since they generally escape different antibodies.

@@

There are caveats that should be kept in mind when using the escape calculator.
First, it only assesses the antigenic effects of mutations, and not how they impact other properties such as viral transmissibility.
Second, the calculator only considers sites in the RBD, and ignores mutations to other regions of spike.
Third, the calculator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
Fourth, the calculator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
Finally, and in our minds most significantly, the calculator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).

Despite all these caveats, the escape calculator yields binding scores that correlate with experimentally measured neutralization titers.
In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape calculator: variants of concern generally have combinations of mutations calculated to have synergistic effects on antigenic escape (eg, 417 and 484) rather than combinations calculated to have redundant effects (eg, 484 and 490).
We suspect the calculator works well because the RBD is the dominant target of neutralizing activity~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high} and the human antibody-response to the early Wuhan-Hu-1 RBD shares broad commonalities across individuals~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.

@@

\subsection{Deep mutational scanning data used by the calculator}
The experimental data used by the escape calculator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955}.
In total, these studies contain data for 36 monoclonal antibodies.
Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the calculator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on sequences reported by Regeneron, Eli Lilly, or their collaborators~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).

The deep mutational scanning measures an escape fraction for each tolerated RBD mutation against each antibody, which represents an estimate of how completely that mutation escapes antibody binding~\citep{greaney2021complete}.
We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions at each site.
The results reported in this paper use the sums as the site-level metric, although the calculator has an option (\texttt{escape\_metric}) to use the mean instead.
We also normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
The results reported in this paper use the normalized site-level metrics, although the calculator has an option (\texttt{escape\_values\_normalized}) to also use the non-normalized values.

>>>>

\lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have emerged only $\sim$2 years since the virus first started to spread in humans.
A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{wang2021antibody,uriu2021neutralization,lucas2021impact}.
Unfortunately, the rate at which new variants arise outstrips the speed at which these experiments can be performed.

@@

\subsection{The escape calculations correlate well with neutralization assays of human polyclonal sera against SARS-CoV-2 variants}
For each set of mutated RBD sites, we can define a quantitative score that represents the polyclonal antibody binding that remains after mutating these sites.
This score is defined using the same principle as the site-wise escape calculator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
This calculation is implemented in the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).

\begin{SCfigure*}[][]
\includegraphics[width=0.7\textwidth]{../results/neut_studies/neut_studies.png}
\caption{Correlation of calculated binding scores with experimentally measured fold changes in neutralization for SARS-CoV-2 variants and mutants (smaller values indicate worse neutralization).
The data of \citet{lucas2021impact} was generated using authentic SARS-CoV-2 and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
The data of \citet{uriu2021neutralization} and \citet{wang2021antibody} was generated using pseudovirus against convalescent (top) or vaccine (bottom) sera, with vaccine sera from Pfizer BNT162b2 or Moderna mRNA-1273 vaccines, respectively. 
The fold changes are geometric means over all subjects in each cohort.
An interactive version of this figure that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/neut_studies.html}.
}
\label{fig:neut_studies}
\end{SCfigure*}

@@

\section{Discussion}
We have described an escape calculator that uses experimental data for 33 monoclonal antibodies to estimate the antigenic effects of arbitrary combinations of mutations to the SARS-CoV-2 RBD.
The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus additive effects on antibody binding.
For instance, sites 417, 484, and 490 are all peaks of antibody escape.
But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have additive effects since they generally escape different antibodies.

@@

There are caveats that should be kept in mind when using the escape calculator.
First, the calculator only considers sites in the RBD, and ignores mutations to other regions of spike.
Second, the calculator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
Third, the calculator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
Finally, and in our minds most significantly, the calculator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).

Despite all these caveats, the escape calculator yields binding scores that correlate with experimentally measured neutralization titers.
In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape calculator: variants of concern generally have combinations of mutations calculated to have additive effects on antibody escape (e.g., 417 and 484) rather than combinations calculated to have redundant effects (e.g., 484 and 490).
We suspect the calculator works well because the RBD is the dominant target of neutralizing activity~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high} and the human antibody-response to the early Wuhan-Hu-1 RBD shares broad commonalities across individuals~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.

@@

\subsection{Deep mutational scanning data used by the calculator}
The experimental data used by the escape calculator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955}.
In total, these studies contain data for 36 monoclonal antibodies.
Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the calculator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on previously reported sequences~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).

The deep mutational scanning measures an escape fraction for each tolerated RBD mutation against each antibody, which represents an estimate of how completely that mutation escapes antibody binding~\citep{greaney2021complete}.
We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions across all tolerated mutations at each site.
The results reported in this paper use the sums as the site-level metric, although the calculator has an option (\texttt{escape\_metric}) to use the mean instead.
We also normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
The results reported in this paper use the normalized site-level metrics, although the calculator has an option (\texttt{escape\_values\_normalized}) to also use the non-normalized values.
